---
title: "pk(pd) assembly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pk(pd) assembly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo=TRUE,
                      error=TRUE,
                      warning=FALSE,
                      message=FALSE)
```

# PK(PD) dataset assembly with apmx
This vignette will demonstrate some of the capabilities of the PK(PD) data assembly functions included in `apmx`. This vignette will use randomly generated source data from an imaginary phase I clinical trial, ABC102.  

## Prepare R workspace and load source data
```{r}
library(dplyr)
library(tidyr)
library(apmx)

EX <- as.data.frame(EX)
PC <- as.data.frame(PC)
DM <- as.data.frame(DM)
LB <- as.data.frame(LB)
```
## Pre-process source datasets
The user often has to pre-process data, even if it is provided as SDTM or ADaM datasets.

### Dose event preparation
We recommend reviewing dose data first, since the doses will tell you a lot about each
subject (dose level, treatment start/end/interruptions, etc.)  
  
ABC102 administers two oral doses to each subject, once at baseline and once on D15. If the source data is well formatted, the user only needs to manually add the compartment.    
```{r}
ex <- EX %>%
  dplyr::mutate(CMT = 1) %>%
  dplyr::select(USUBJID, STUDYID, EXSTDTC, VISIT, EXSTDY, EXTPTNUM, EXDOSE,
                CMT, EXTRT, TPTC, EXROUTE, EXDOSFRQ, EXDOSU)
```

### PK event preparation
Next, prepare the PK observation data.  
  
Some SDTM and ADaM datasets do not have the information we need in the ideal format. In this example, the user manually creates numeric time points.  
```{r}
pc <- PC %>%
  dplyr::filter(PCSTAT=="Y") %>%
  dplyr::mutate(CMT = 2,
                TPT = dplyr::case_when(PCTPT=="<1 hour Pre-dose" ~ 0,
                                       PCTPT=="30 minutes post-dose" ~ 0.5/24,
                                       PCTPT=="1 hour post-dose" ~ 1/24,
                                       PCTPT=="2 hours post-dose" ~ 2/24,
                                       PCTPT=="4 hours post-dose" ~ 4/24,
                                       PCTPT=="6 hours post-dose" ~ 6/24,
                                       PCTPT=="8 hours post-dose" ~ 8/24,
                                       PCTPT=="12 hours post-dose" ~ 12/24,
                                       PCTPT=="24 hours post-dose" ~ 24/24,
                                       PCTPT=="48 hours post-dose" ~ 48/24)) %>%
  dplyr::select(USUBJID, PCDTC, PCDY, VISIT, TPT, PCSTRESN,
                PCLLOQ, CMT, PCTEST, PCTPT, PCSTRESU)
```

## Create PK dataset
We now have all of the information we need to build a PK dataset with `pk_build()`  
```{r}
pkdf <- apmx::pk_build(ex = ex,
                       pc = pc,
                       time.rnd = 3)
```

`pk_build()` includes additional parameters to help format the dataset.  
  
The `cycle.length` and `time.rnd` parameters will make the time variables easier to read.  
```{r}
pkdf <- apmx::pk_build(ex = ex,
                       pc = pc,
                       time.units = "days",
                       cycle.length = 14,
                       time.rnd = 3)
```

## Add covariates
The `cov_apply()` function will add additional covariates from a source data frame.  
  
In this example, we will add demographic covariates to the pk dataset. AGE is continuous,
but SEX, RACE, and ETHNIC are categorical. The function will convert AGE to BAGE (baseline age) and convert SEX, RACE, and ETHNIC to NSEXC, NRACEC, and NETHNICC (character label) and NSEX, NRACE, NETHNIC (encoded numeric).  
```{r}
dm <- DM %>%
  dplyr::select(USUBJID, AGE, SEX, RACE, ETHNIC)

pkdf <- apmx::pk_build(ex = ex,
                       pc = pc,
                       cycle.length = 14,
                       time.rnd = 3) %>%
  apmx::cov_apply(dm)
```

Subject-level covariates can also be added in the `pk_build()` function.  
  
We can also use the `cov_find()` function to determine which covariates are included in the dataset.  
```{r}
pkdf <- apmx::pk_build(ex = ex,
                       pc = pc,
                       sl.cov = dm,
                       cycle.length = 14,
                       time.rnd = 3)

apmx::cov_find(pkdf, cov="categorical", type="numeric")
apmx::cov_find(pkdf, cov="categorical", type="character")
apmx::cov_find(pkdf, cov="continuous", type="numeric")
```

Time-varying covariates can also be added to a dataset with either `cov_apply()` or `pk_build()`.  
```{r, warning=T}
lb <- LB %>%
  dplyr::filter(LBCOMPFL=="Y") %>%
  dplyr::filter(LBPARAMCD=="ALB") %>%
  dplyr::mutate(ALB = as.numeric(LBORRES),
                DTIM = LBDT) %>%
  dplyr::select(USUBJID, DTIM, ALB)

pkdf <- apmx::pk_build(ex = ex,
                       pc = pc,
                       cycle.length = 14,
                       time.rnd = 3) %>%
  apmx::cov_apply(dm) %>%
  apmx::cov_apply(lb,
                  time.by = "DTIM",
                  cov.rnd = 1)

pkdf <- apmx::pk_build(ex = ex,
                       pc = pc,
                       sl.cov = dm,
                       tv.cov = lb,
                       cycle.length = 14,
                       time.rnd = 3,
                       cov.rnd = 1)
```

## Warnings and Errors
You will notice that `pk_build()` throws a warning after adding in the time-varying
covariates that a subject is missing ATFD. This is because we added albumin values
for a subject who does not have dose or pk events in the dataset. That record is
automatically commented out of the dataset.  
  
There are other error messages and warnings `pk_build()` writes to help diagnose your
dataset. We will force a few of them. Let's start by adding covariates for some, but not all, subjects.  
```{r, warning=T}
nrow(dm)
dm1 <- dm[1:10,] #drop several rows of covariates

pkdf <- apmx::pk_build(ex = ex,
                       pc = pc,
                       sl.cov = dm1,
                       cycle.length = 14,
                       time.rnd = 3,
                       cov.rnd = 1)
```
This warning lets you know there are subjects with dose or obsrevation events who do not have any covariates. All missing covariates values are filled with the `na` parameter, which
is set to -999 by default.  
  
Next, let's create a dataset where the date/time for pk events are missing time.  
```{r, warning=T}
pc1 <- pc %>%
  dplyr::mutate(PCDTC = substr(PCDTC, 1, 10))

pkdf <- apmx::pk_build(ex = ex,
                       pc = pc1,
                       sl.cov = dm,
                       cycle.length = 14,
                       time.rnd = 3,
                       cov.rnd = 1)
```
This error message will not let you build a dataset with incorrectly formatted date/time. You should review the date/time data for all rows to ensure you are not missing any information.  
  
Next, let's alter one of the nominal days for a pc draw so it is incorrect.  
```{r, warning=T}
pc1 <- pc
pc1$TPT[5] <- 0.05 #change a nominal timepoint to an incorrect time

pkdf <- apmx::pk_build(ex = ex,
                       pc = pc1,
                       sl.cov = dm,
                       cycle.length = 14,
                       time.rnd = 3,
                       cov.rnd = 1)
```
This error message let's you know that events occurred out of protocol order. This is indicated by the nominal times, which are not increasing at row 6.  
  
Next, let's remove one of the date/times from the ex domain to make it missing.  
```{r, warning=T}
ex1 <- ex
ex1$EXSTDTC[1] <- NA #remove date/time for a subject's dose

pkdf <- apmx::pk_build(ex = ex1,
                       pc = pc,
                       sl.cov = dm,
                       cycle.length = 14,
                       time.rnd = 3,
                       cov.rnd = 1)
```
This warning will let you know the subject(s) that have events without ATFD (actual time since first dose), usually due to missing DTIM or missing time since first dose. These comments are automatically commented out of the dataset with the C flag.  
  
`pk_build()` can impute missing times with two different methods.  
  
  * Method 1 will impute the nominal time as the actual time. This method is best used for preclinical projects or projects with no actual times at all.  
  * Method 2 will estimate the actual time based on other events that occurred that day. This method is best used for phase I-III projects where only some date/times are missing.  
```{r, warning=T}
pkdf <- pk_build(ex = ex1,
                 pc = pc,
                 sl.cov = dm,
                 cycle.length = 14,
                 time.rnd = 3,
                 cov.rnd = 1,
                 impute = 2)
```
The impute method 2 successfully estimates the time of the missing dose and the warning disappears. All automatic imputations are marked with the "IMPEX", "IMPFEX", and "IMPDV" flags.
